sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-22 doc/paper.Rmd
Substantial paper editing

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], Other authors\n| $^1$Pacific Biological Station Fisheries and Oceans Canada, \n| Nanaimo, BC, Canada\n|
  $^2$Other addresses\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs). These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models. Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB). We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the package to related software.
    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
    | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, Template Model Builder
---

@@

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].
Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data.
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle1998; @chiles1999].

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

@@

sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates. 
The data can be collected across discrete points in time.
Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR).
sdmTMB can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram)).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit).
See Appendix TODO for a full description of the statistical model.

@@

The GLMMs underpinning sdmTMB models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim$ 0.13 (REF)---can be shared or estimated separately.

@@

By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix TODO).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix TODO).

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

@@

Here, `cutoff` defines the minimum allowed distance between points in the units of `X` and `Y` (km).
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but increase the accuracy of the SPDE approximation.
The circle area corresponds to density of Pacific Cod for individual trawls.
The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
These predictions are made omitting the spatial random field.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
(D) Conditional effect of depth modelled as a penalized smoother.
(E) Standardized area-weighted population index derived from a spatiotemporal version of this model.

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution. 
We fit counts observed in North America during annual Christmas Bird Counts (@cbc) using a negative binomial distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3):

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. 
We provide several approaches to assist this process: 
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
<!-- We allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]; however, because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation. -->
<!-- The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. -->
<!-- The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model. -->
<!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection (e.g., via expected log predictive density, ELPD, REF) or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`). 
(4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
Simulation-based residuals [@DHARMa], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
(5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
<!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->

@@

knitr::kable(d, format = "latex", caption = "Comparison of functionality between seven R packages that can fit geostatistical GLMMs.", 
  booktabs = TRUE, 
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>% 
  kableExtra::footnote(
    number = c(
      "Technically possible but non-trivial",
      "Penalized smoother GAMs",
      "Spatiotemporal fields as random walk only",
      "SPDE approach as in Miller et al. (2019)",
      "Does have predictive process knots",
      "Zero-inflated NB2 only",
      "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}",
      "Technically possible but non-trivial",
      "Technically possible but non-trivial",
      "Available in a companion package INLAutils"
    ),
    escape = FALSE
  )
```

@@

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).
<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
We provide examples of the syntax and the reproducibility of results achieved when similar models are fit using INLA (Appendix 4) and VAST (Appendix 5).
<!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
<!-- TODO: mention inlabru here -->

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Appendix 6 TODO, Fig. \@ref(fig:timing)).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data points slightly shift the lines up without affecting the overall pattern substantially (not shown).
Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speedup for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

@@

None?

@@

>>>>

---
title: "sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca], 
  Eric J. Ward$^2$, 
  Philina A. English$^1$, 
  Lewis A. K. Barnett$^3$\n| 
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Analyzing geostatistical data---spatially referenced observations---is increasingly common in ecology for applications such as species distribution models (SDMs). These data present statistical challenges to modelling them effectively and efficiently; appropriate models such as GLMMs (generalized linear mixed effects models) with Gaussian random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2: We introduce the R package sdmTMB, which implements predictive-process SPDE-based (stochastic partial differential equation) spatial and spatiotemporal models. Estimation is performed with maximum marginal likelihood with Template Model Builder (TMB). We describe the basic functionality of sdmTMB and the statistical model, illustrate its use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3:  We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of glm(), lme4, glmmTMB, or mgcv, (2) estimation is often faster than alternatives, (3) sdmTMB allows for simple out-of-sample model validation, (4) non-stationary processes are easily constructed with a formula interface (time-varying coefficients, spatially varying coefficients [SVC]), and (5) sdmTMB includes features not available as a combination in related packages (penalized smoothers and break-point effects, anisotropy, population index standardization).
    | 4: We hope that user-friendly interfaces such as sdmTMB open this useful class of models to a wider audience of users.
    | Keywords: Gaussian Markov random fields, generalized linear mixed effects models, INLA, species distribution modelling, R package, Template Model Builder
---

@@

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions compared to ignoring it [e.g., @shelton2014], and can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

Geostatistical GLMMs (generalized linear mixed effects models) with spatially correlated random effects are a class of models appropriate for these data [@rue2005gmrf; @diggle2007; @cressie2011].
Similarly to how random intercepts can account for correlation among groups, spatial or spatiotemporal random effects can account for unmeasured variables causing observations to be correlated in space or space and time.
A common approach to modelling these spatial effects is with Gaussian random fields (GRFs), where the random effects describing the spatial patterning are assumed drawn from a multivariate normal (MVN) distribution, constrained by some covariance function such as the exponential or Matérn [@cressie1993; @diggle2007; @chiles1999].

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

(ref:diagram-fig) Components of an sdmTMB model with illustrations, descriptions, examples, notation, and example code.
An sdmTMB model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix 1 for a full description of the model.
Note that `s()` denotes a smoother as in mgcv [@wood2017a], `breakpt()` denotes a breakpoint 'hockey-stick' shape [e.g., @barrowman2000], `(1|g) ` denotes a random intercept by group `g`, and `~ 0` is used in an R formula to omit an intercept.

@@

sdmTMB fits GLMMs to spatial or spatiotemporal geostatistical data.
Geostatistical data simply refers to data observed at specific spatial coordinates reflecting some underlying spatial process [@rossi1992; @diggle2007]. 
These data can be collected across discrete points in time.
Areal data (data on polygons or grids) may be analyzed using other spatial models, including conditional autoregressive models (CAR) [e.g., @verhoef2018].
sdmTMB can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is the parameters describing spatial covariance can be estimated (e.g., spatial variance and range---the distance at which points are effectively independent).

The process component of an sdmTMB model can be formed by any combination of main ("fixed") effects, spatial intercept random fields, spatiotemporal intercept random fields, IID random intercepts, time-varying effects, and spatially varying effects (Fig. \@ref(fig:diagram), Appendix 1).
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].

@@

The GLMMs underpinning sdmTMB models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], most smooths in mgcv [@wood2017a]).
The random fields in sdmTMB are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately.

@@

By default, if spatiotemporal fields are included, they are assumed independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram); Appendix 1).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey-stick models, `breakpt()`, @barrowman2000] or logistic functions, `logistic()` (Appendix 1).

The sdmTMB model is fit by maximum marginal likelihood.
Internally, a TMB [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an sdmTMB model to the R package tmbstan [@monnahan2018] to estimate the joint posterior distribution.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y` (km).
Alternatively, we could have created any mesh via the INLA package and supplied it to `make_mesh()`. 
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

(ref:pcod-fig-cap) Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points).
Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation.
The circle area corresponds to density of Pacific Cod for individual trawls.
The line represents the estimate and the ribbon indicates a 95\% confidence interval (+/- 2 SEs).
These predictions are made omitting the spatial random field.
(B) Spatial random field:
these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect.
(C) Overall prediction: these estimates represent the combination of all fixed and random effects.
(D) Conditional effect of depth modelled as a penalized smoother.
(E) Standardized area-weighted population index derived from a spatiotemporal version of this model.

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter. 
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution. 
We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO (Appendix 3).

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. 
We provide several approaches to assist this process: 
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
<!-- We allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]; however, because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation. -->
<!-- The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. -->
<!-- The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model. -->
<!-- Validation of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation: -->
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or evaluate residuals (see `?extract_mcmc`). 
(4) The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
Simulation-based residuals [@DHARMa], one-step-ahead residuals [@thygesen2017a], or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are slower.
(5) The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
<!-- Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE). -->
<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->

@@

knitr::kable(d, format = "latex", caption = "Comparison of functionality between seven R packages that can fit geostatistical GLMMs.", 
  booktabs = TRUE, 
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>% 
  kableExtra::footnote(
    number = c(
      "Technically possible but non-trivial",
      "Penalized smoother GAMs",
      "Spatiotemporal fields as random walk only",
      "SPDE approach as in Miller et al. (2019)",
      "Does have predictive process knots",
      "Zero-inflated NB2 only",
      "Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}",
      "Technically possible but non-trivial",
      "Technically possible but non-trivial"
    ),
    escape = FALSE
  )
```

@@

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).
<!-- TODO: add something about the results of the INLA comparison (Appendix 4) and the VAST comparison (Appendix 5) -->
We provide comparisons of the syntax and the reproducibility of results from models fit using INLA (Appendix 4) and VAST (Appendix 5).
<!-- We ran into problems with AR1 spatiotemporal fields in inlabru, otherwise we would have included both INLA and inlabru syntax.
<!-- TODO: mention inlabru here -->

@@

We ran a simple speed comparison between sdmTMB, inlabru, and mgcv for fitting a SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix 6).
In this test, for mesh resolutions under 500 nodes, sdmTMB fit the models the fastest.
inlabru was the least affected by mesh resolution and beyond 500 nodes was faster than sdmTMB with default sdmTMB settings.
mgcv was slightly slower than sdmTMB and was more affected by mesh resolution.
With the random effect normalizing constant calculated in R (`normalize = TRUE`), sdmTMB was the fastest across all mesh resolutions (this option may not be possible or as fast for all models).
Results with up to an order of magnitude more data slightly shift the lines up without affecting the overall pattern substantially (not shown).
Our test was with one core and default R algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for sdmTMB and inlabru and a ~7--9-fold speed increase for mgcv.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->

@@

TODO: None?

@@

TODO\
Will be on CRAN\
Paper source code will be on Zenodo\
<https://github.com/pbs-assess/sdmTMB>\
<https://github.com/seananderson/sdmTMB-paper>
